---
title: "Experiment 2 -- Red Eggs"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Experiment 2 Red Eggs--Overview    

As mentioned in the previous document, the red egg stage is the second of three
developmental stages at which egg viability was assessed. In normal development,
there is a near total correlation between egg viability and eggs turning from
white to red. Irradiation, however, can cause fatal developmenal problems 
aftere this color shift which reduced the reliablity of egg color as an 
indicator of fertility. As previously, the sexes are separated because of 
expected differences in radiosensitivy and the proportion of red eggs are 
by x-ray treatment (0, 100, 200, 300, or 400 Gy)

```{r load_data}
library(readxl)
# Load data and display first few rows
df_red <- read_excel("Dataset.xlsx", sheet = "Development_400Gy")

# Set x-ray dose as a factor (categorical)
df_red$dose <- as.factor(df_red$dose)
```

Separate sexes

```{r fem_dat, message=FALSE, warning=FALSE}
library(dplyr)
# Isolate data for females and display the first few rows
females <- df_red %>% 
  filter(sex == "female")
females
```

```{r male_dat, message=FALSE, warning=FALSE}
# Isolate data for males and display the first few rows
males <- df_red %>% 
  filter(sex == "male")
males
```


## Summary and statistical analysis of red eggs from irradiated females

```{r fem_trt_means, message=FALSE, warning=FALSE}
library(FSA)            # for SE
# n, mean, and SE by dose for females
females %>% 
  group_by(dose) %>%
  summarise(nObs = sum(!is.na(sum_red)),
            mn = mean(sum_red, na.rm = TRUE),
            sem = se(sum_red),
            pct_sterile = 100*(1-(sum(sum_red > 0)/nObs)))
```

The mean and SE are consistent with overdispersed count data. The negative 
binomial is appropriate for such data. The percent sterility calculation
reminds us that some females are laying at least some red eggs.

```{r females_nb, message=FALSE, warning=FALSE}
# Analysis of Deviance using GLM with negative binomial--females
library(MASS)
m1 <- MASS::glm.nb(sum_red ~ dose, data = females)
summary(m1)
```
Examine model fit for females

```{r fem_model_fit,  message=FALSE, warning=FALSE}
# Calculate p-value under the chi-squared distribution
dev <- deviance(m1)
df <- df.residual(m1)
p_value <- pchisq(dev, df, lower.tail = FALSE)
cat("Chi-square goodness-of-fit p-value:", p_value, "\n")
```
The residual deviance is not significantly different (P > 0.05) from the 
expected value for the residual degrees of freedom.

```{r female_type2, message=FALSE, warning=FALSE}
library(car)
# Model parameters with type II degrees of freedom
Anova(m1)
```
Overall, there were significant differences. Proceeding to comparisons of 
multiple means.


```{r female_meanssep, message=FALSE, warning=FALSE}
library(emmeans)
library(multcomp)
# Means separation
emm_fem <- emmeans(m1, ~ dose, type = "response")
multcomp::cld(emm_fem, Letters = LETTERS, decreasing = TRUE)
```

## Summary and statistical analysis of red eggs from females mated with irradiated males

```{r male_trt_means}
# n, mean, and SE by dose for females
males %>% 
  group_by(dose) %>%
  summarise(nObs = sum(!is.na(sum_red)),
            mn = mean(sum_red, na.rm = TRUE),
            sem = se(sum_red),
            pct_sterile = 100*(1-(sum(sum_red > 0)/nObs)))
```

```{r males_nb, message=FALSE, warning=FALSE}
# Analysis of Deviance using GLM with negative binomial--males
m2 <- MASS::glm.nb(sum_red ~ dose, data = males)
summary(m2)
```

Examine model fit for males

```{r male_model_fit,  message=FALSE, warning=FALSE}
# Calculate p-value under the chi-squared distribution
dev <- deviance(m2)
df <- df.residual(m2)
p_value <- pchisq(dev, df, lower.tail = FALSE)
cat("Chi-square goodness-of-fit p-value:", p_value, "\n")
```
The residual deviance is not significantly different (P > 0.05) from the 
expected value for the residual degrees of freedom.

```{r male_type2, message=FALSE, warning=FALSE}
# Model parameters with type II degrees of freedom
Anova(m2)
```

Overall, there were significant differences. Proceeding to comparisons of 
multiple means.


```{r male_meanssep, message=FALSE, warning=FALSE}
# Model parameters with type II degrees of freedom
emm_males <- emmeans(m2, ~ dose, type = "response")
multcomp::cld(emm_males, Letters = LETTERS, decreasing = TRUE)
```

